Meta-Analysis of Clinical Phenotype and Patient Survival in Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies Due to Bi-allelic Loss of Function Variants in SMPD

### 2022-09-15 paper/paper.tex
simplified

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  authoryear,
  preprint,
  3p]{elsarticle}

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{caption}
\usepackage{longtable}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Clinical Phenotype and Survival Analysis of Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies Due to Bi-allelic Loss of Function Variants in SMPD4},
  pdfauthor={Dean Marchiori},
  pdfkeywords={NEDMABA, SMPD4},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\setlength{\parindent}{6pt}
\begin{document}

\begin{frontmatter}
\title{Clinical Phenotype and Survival Analysis of Neurodevelopmental
Disorder with Microcephaly, Arthrogryposis, and Structural Brain
Anomalies Due to Bi-allelic Loss of Function Variants in \emph{SMPD4}}
\author[]{Dean Marchiori%
\corref{cor1}%
}
 \ead{deanmarchiori@gmail.com}

\cortext[cor1]{Corresponding author}

\begin{abstract}
A recently described, rare genetic condition known as Neurodevelopmental
Disorder with Microcephaly, Arthrogryposis, and Structural Brain
Anomalies (NEDMABA) has been identified in children with bi-allelic
loss-of-function variants in \emph{SMPD4}. The progression of this
condition is not well understood with the limited case reports described
so far exhibiting a severe and clinically diverse phenotype. A gap
exists in the understanding of associations present in the heterogenous
features of the clinical phenotype, and the expected survival
probabilities of affected individuals. This is driven in part to the
paucity of analysis-ready data on reported cases. This analysis aims to
collate and standardise available case reports into a common dataset, to
analyse and identify meaningful clusters in the clinical phenotype, and
to quantify the survival probability for children with NEDMABA. To
overcome the challenge of multidimensional data on very few subjects, we
employ Multiple Correspondence Analysis (MCA) as a dimension reduction
technique, which is then subject to cluster analysis and interpretation.
To account for censoring in the data, Kaplan-Meier estimation is
formulated to calculate patient survival time. The analysis correctly
detected the classic phenotype for this condition, as well as a new
distinct feature-cluster relating to findings of vocal cord paralysis,
feeding dysfunction and respiratory failure. The survival probability
for those affected was found to decline sharply in early infancy with
median survival of 150 days, but with some surviving as long as 12.5
years. This wide range of outcomes is provisionally associated with
different variant types however this conclusion could not be validated
based on very low sample sizes. An \texttt{R} package called
\texttt{SMPD4} was developed to publish standardised analysis-ready
datasets used in this study. This analysis represents the first of its
kind to help describe associations and trajectories of individuals with
this newly reported condition, despite challenges with sparse and
inconsistent data. This analysis can provide clinicians and genetic
counsellors with better information to aide in decision making and
support for families with this rare condition.
\end{abstract}

\begin{keyword}
    NEDMABA \sep 
    SMPD4
\end{keyword}
\end{frontmatter}\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, frame hidden, boxrule=0pt, enhanced, interior hidden, breakable, sharp corners]}{\end{tcolorbox}}\fi

@@

Neurodevelopment disorders are a diverse and heterogeneous group of
conditions impacting the development of the nervous system, brain
function, physical development and learning ability. A recently
identified condition known as Neurodevelopmental Disorder with
Microcephaly, Arthrogryposis, and Structural Brain Anomalies (NEDMABA)
(MIM:\href{https://omim.org/entry/618622}{618622}) has been described in
children with bi-allelic loss-of-function variants in \emph{SMPD4}
\citep{magini2019loss}. Sphingomyelinases such as \emph{SMPD4} play an
important cellular role by hydrolyzing sphingomyelin into ceramide and
phosphorylcholine. \emph{SMPD4} specifically encodes one of 4 neutral
sphinogmyelinases, nSMase3 (MIM:
\href{http://omim.org/entry/610457}{610457}). So far, less than 50 cases
of this rare disorder have been reported in the literature. Clinical
phenotype data of these cases is largely heterogeneous with severe
neurological complications and early demise a key feature. This makes
the identification, diagnosis and ongoing management of cases a
challenge for patients, their families and medical practitioners.

A detailed study by \citet{magini2019loss} involved 12 unrelated
families with 32 individuals (21 with detailed clinical information).
Presentations were of microcephaly, simplified gyral pattern of the
cortex, hypomyelination, cerebellar hypoplasia, congenital
arthrogryposis, and early fetal/postnatal demise. Despite this being the
largest cohort studied, the clinical features and survival times among
participants varied greatly. Three missense changes were noted in the
study with affected children in these families often showing a milder
presentation suggestive of possible residual function. In these cases
individuals were able to develop independent motor skills, have mild
intellectual disability and arthrogryposis without evidence of
simplified gyral patterns on brain MRI. Other patients with truncating
variants are shown to have more severe presentations, while a range of
additional significant clinical features were reported involving
dysmorphic facial features, seizure, vocal cord paralysis and hearing
impairment. \citet{ravenscroft2021neurogenetic} studied a family from
Melbourne exhibiting features involving arthrogryposis multiplex
congenita, complex brain malformations, small for gestation age and
hypoplasia of the corpus callosum. In two of the three related cases
were additional features of microcephaly, congenital encephalopathy,
cerebellar malformation and hypoplasia and hypomyelination. A case study
from China is described in \citet{ji2022case} involving a girl
presenting in infancy with intrauterine growth restriction,
microcephaly, postnatal developmental delay, arthrogryposis,
hypertonicity, seizure, and hypomyelination on brain magnetic resonance
imaging. The authors report parallels with two cases showing the same
homozygous null variant. An individual reported in
\citet{monies2019lessons} presented with distinct symptoms of brain
atrophy and skeletal dysplasia whereas the case in
\citet{magini2019loss} with the same variant exhibited more typical
clinical features. A recent study by \citet{bijarnia2022growth} presents
the case of a 22-month old girl presenting with the typical phenotype of
neurodevelopmental delay, prenatal onset growth failure, arthrogryposis,
microcephaly and brain anomalies including severe hypomyelination,
simplified gyral pattern and hypoplasia of corpus callosum and
brainstem. Notably, there was also additional non-typical clinical
findings of nystagmus and visual impairment secondary to macular
dystrophy and retinal pigment epithelial stippling at posterior pole.

Further work to collate and analyse data relating to NEDMABA is
challenging. While \citet{magini2019loss} included a supplementary
clinical phenotype data set in their study, these data are not suitable
for statistical analysis. Many of the clinical features are represented
as text descriptions and a variety of non-standard terminologies are
applied between cases, making machine interpretation of the data
difficult. Other studies \citep{ji2022case, bijarnia2022growth} present
only written case reports with some tabulated summaries. While larger
studies \citep{ravenscroft2021neurogenetic, monies2019lessons} focus
more on genetic data, with less detail included on the clinical
presentation of individuals. To overcome these data challenges,
\citet{JSSv059i10} proposes a data structure know as `tidy' data which
organises each observation in a row, with each feature as its own column
and each value in just one cell. In this case detailed text-based data
can be transformed to high dimensional binary indicators of key clinical
features. This structure is conducive to effective statistical analysis
and integrates deliberately with the \emph{tidyverse} \citep{tidyverse}
collection of packages for data analysis in R \citep{rbase}.

Further analysis to better describe rare genetic conditions was explored
in \citet{diaz2020phenotype} in the context of large scale
genotype-phenotype analysis. The authors argue patients with rare
disorders (often in small samples) frequently present with varied
symptoms that do not match exactly with the described phenotype. When
considering the situation of low sample size data with many features,
methods in the field of multivariate statistical analysis are commonly
deployed. Here the aim is to find substructure in the data, or a simple
representation of the multi-dimensional space. Methods, such as Multiple
Correspondence Analysis \citep{le2010multiple} are appropriate for high
dimensional data comprised of binary indicators. This technique is cited
in many studies in the application of dimension reduction and clustering
of comorbidities and phenotypes from disease
\citep{han2018cluster, costa2013use}.

In terms of patient survival, methods in statistical survival analysis
are well suited and commonly used in the field of genomics
\citep{chen2014survival}. These methods in particular provide a
mechanism for dealing with right-censoring, a common feature of clinical
studies where the clinical outcome of interest may not be known by the
end of the study period.

An open research question exists around the clinical pathway and
survival time of children exhibiting bi-allelic loss-of-function
variants in \emph{SMPD4}. With current research highlighting a diverse
and heterogeneous phenotype, the relationship between these diverse
features is not well understood. Furthermore, the survival time of
affected children has not been extensively analysed despite early-demise
featuring as a severe outcome. While a connection has been highlighted
between some missense variants and a milder presentation with longer
survival \citep{magini2019loss}, this hypothesis has not been analysed
further.

@@

The data contained in \citet{magini2019loss} ``Summary of SMPD4-Related
Clinical Phenotype'' is an excel spreadsheet tabulating each of the 21
individuals from the study with clinical details recorded. The file has
21 columns (one for each individual), and 64 rows (one for each
phenotype or clinical remark).

The data were read into \texttt{R} \citep{rbase} without any changes, so
as to preserve the reproducibility of the data transformation steps. The
data were transposed to ensure it could be presented as \emph{tidy}
formatted data \citep{JSSv059i10}. This requires that each variable
(clinical phenotype) forms a column; each observation (individual) forms
a row and each type of observational unit forms a table (every cell has
just one item). This resulted in a long-formatted data set of 21 rows
(individuals) and 64 columns (clinical features). In many cases, key
clinical information was entered as free-text descriptions, which
rendered any attempt of meaningful analysis impractical
(Table~\ref{tbl-nontidy}). In these cases, the text was tokenised by
separating the list of clinical observations at each comma and forming a
binary indicator column noting its presence `1' or absence `0'
(Table~\ref{tbl-tidy}).

\hypertarget{tbl-nontidy}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2083}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.5139}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2778}}@{}}
\caption{\label{tbl-nontidy}Example of non-tidy free-text descriptions
of features}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Family 1- Individual 1
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Family 1- Individual 4
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Family 1- Individual 1
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Family 1- Individual 4
\end{minipage} \\
\midrule()
\endhead
Facial dysmorphisms & short palpebral fissures, large ears, simple
helices, smooth philtrum, thin lips, bilateral simian creases & short
palpebral fissures, receding forehead, thin upper lip \\
\bottomrule()
\end{longtable}

\hypertarget{tbl-tidy}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1644}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1781}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1644}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1644}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1644}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1644}}@{}}
\caption{\label{tbl-tidy}Example of tidy formatted data where
individuals are transposed into rows and text into binary
indicators}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
id
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
short palpebral fissures
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
large ears
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
simple helices
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
receding forehead
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\ldots{}
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
id
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
short palpebral fissures
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
large ears
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
simple helices
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
receding forehead
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\ldots{}
\end{minipage} \\
\midrule()
\endhead
Family 1- Individual 1 & 1 & 1 & 1 & 0 & \ldots{} \\
Family 1- Individual 4 & 1 & 0 & 0 & 1 & \ldots{} \\
\bottomrule()
\end{longtable}

@@

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{caption}
\usepackage{longtable}
\KOMAoption{captions}{tableheading}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Meta-Analysis of Clinical Phenotype and Patient Survival in Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies Due to Bi-allelic Loss of Function Variants in SMPD},
  pdfauthor={Dean Marchiori},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Meta-Analysis of Clinical Phenotype and Patient Survival in
Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and
Structural Brain Anomalies Due to Bi-allelic Loss of Function Variants
in SMPD}
\author{Dean Marchiori}
\date{9/15/22}

\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, frame hidden, sharp corners, boxrule=0pt, interior hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi

\hypertarget{abstract}{%
\section{Abstract}\label{abstract}}

A recently described, rare genetic condition known as Neurodevelopmental
Disorder with Microcephaly, Arthrogryposis, and Structural Brain
Anomalies (NEDMABA) has been identified in children with bi-allelic
loss-of-function variants in \emph{SMPD4}. The progression of this
condition is not well understood with the limited case reports described
so far exhibiting a severe and clinically diverse phenotype. A gap
exists in the understanding of associations present in the heterogenous
features of the clinical phenotype, and the expected survival
probabilities of affected individuals. This is driven in part to the
paucity of analysis-ready data on reported cases. This analysis aims to
collate and standardise available case reports into a common dataset, to
analyse and identify meaningful clusters in the clinical phenotype, and
to quantify the survival probability for children with NEDMABA. To
overcome the challenge of multidimensional data on very few subjects, we
employ Multiple Correspondence Analysis (MCA) as a dimension reduction
technique, which is then subject to cluster analysis and interpretation.
To account for censoring in the data, Kaplan-Meier estimation is
formulated to calculate patient survival time. The analysis correctly
detected the classic phenotype for this condition, as well as a new
distinct feature-cluster relating to findings of vocal cord paralysis,
feeding dysfunction and respiratory failure. The survival probability
for those affected was found to decline sharply in early infancy with
median survival of 150 days, but with some surviving as long as 12.5
years. This wide range of outcomes is provisionally associated with
different variant types however this conclusion could not be validated
based on very low sample sizes. An \texttt{R} package called
\texttt{SMPD4} was developed to publish standardised analysis-ready
datasets used in this study. This analysis represents the first of its
kind to help describe associations and trajectories of individuals with
this newly reported condition, despite challenges with sparse and
inconsistent data. This analysis can provide clinicians and genetic
counsellors with better information to aide in decision making and
support for families with this rare condition.

@@

Neurodevelopment disorders are a diverse and heterogeneous group of
conditions impacting the development of the nervous system, brain
function, physical development and learning ability. A recently
identified condition known as Neurodevelopmental Disorder with
Microcephaly, Arthrogryposis, and Structural Brain Anomalies (NEDMABA)
(MIM:\href{https://omim.org/entry/618622}{618622}) has been described in
children with bi-allelic loss-of-function variants in \emph{SMPD4}
(Magini et al. 2019). Sphingomyelinases such as \emph{SMPD4} play an
important cellular role by hydrolyzing sphingomyelin into ceramide and
phosphorylcholine. \emph{SMPD4} specifically encodes one of 4 neutral
sphinogmyelinases, nSMase3 (MIM:
\href{http://omim.org/entry/610457}{610457}). So far, less than 50 cases
of this rare disorder have been reported in the literature. Clinical
phenotype data of these cases is largely heterogeneous with severe
neurological complications and early demise a key feature. This makes
the identification, diagnosis and ongoing management of cases a
challenge for patients, their families and medical practitioners.

A detailed study by Magini et al. (2019) involved 12 unrelated families
with 32 individuals (21 with detailed clinical information).
Presentations were of microcephaly, simplified gyral pattern of the
cortex, hypomyelination, cerebellar hypoplasia, congenital
arthrogryposis, and early fetal/postnatal demise. Despite this being the
largest cohort studied, the clinical features and survival times among
participants varied greatly. Three missense changes were noted in the
study with affected children in these families often showing a milder
presentation suggestive of possible residual function. In these cases
individuals were able to develop independent motor skills, have mild
intellectual disability and arthrogryposis without evidence of
simplified gyral patterns on brain MRI. Other patients with truncating
variants are shown to have more severe presentations, while a range of
additional significant clinical features were reported involving
dysmorphic facial features, seizure, vocal cord paralysis and hearing
impairment. Ravenscroft et al. (2021) studied a family from Melbourne
exhibiting features involving arthrogryposis multiplex congenita,
complex brain malformations, small for gestation age and hypoplasia of
the corpus callosum. In two of the three related cases were additional
features of microcephaly, congenital encephalopathy, cerebellar
malformation and hypoplasia and hypomyelination. A case study from China
is described in Ji et al. (2022) involving a girl presenting in infancy
with intrauterine growth restriction, microcephaly, postnatal
developmental delay, arthrogryposis, hypertonicity, seizure, and
hypomyelination on brain magnetic resonance imaging. The authors report
parallels with two cases showing the same homozygous null variant. An
individual reported in Monies et al. (2019) presented with distinct
symptoms of brain atrophy and skeletal dysplasia whereas the case in
Magini et al. (2019) with the same variant exhibited more typical
clinical features. A recent study by Bijarnia-Mahay et al. (2022)
presents the case of a 22-month old girl presenting with the typical
phenotype of neurodevelopmental delay, prenatal onset growth failure,
arthrogryposis, microcephaly and brain anomalies including severe
hypomyelination, simplified gyral pattern and hypoplasia of corpus
callosum and brainstem. Notably, there was also additional non-typical
clinical findings of nystagmus and visual impairment secondary to
macular dystrophy and retinal pigment epithelial stippling at posterior
pole.

Further work to collate and analyse data relating to NEDMABA is
challenging. While Magini et al. (2019) included a supplementary
clinical phenotype data set in their study, these data are not suitable
for statistical analysis. Many of the clinical features are represented
as text descriptions and a variety of non-standard terminologies are
applied between cases, making machine interpretation of the data
difficult. Other studies (Ji et al. 2022; Bijarnia-Mahay et al. 2022)
present only written case reports with some tabulated summaries. While
larger studies (Ravenscroft et al. 2021; Monies et al. 2019) focus more
on genetic data, with less detail included on the clinical presentation
of individuals. To overcome these data challenges, Wickham (2014)
proposes a data structure know as `tidy' data which organises each
observation in a row, with each feature as its own column and each value
in just one cell. In this case detailed text-based data can be
transformed to high dimensional binary indicators of key clinical
features. This structure is conducive to effective statistical analysis
and integrates deliberately with the \emph{tidyverse} (Wickham et al.
2019) collection of packages for data analysis in R (R Core Team 2022).

Further analysis to better describe rare genetic conditions was explored
in Dı́az-Santiago et al. (2020) in the context of large scale
genotype-phenotype analysis. The authors argue patients with rare
disorders (often in small samples) frequently present with varied
symptoms that do not match exactly with the described phenotype. When
considering the situation of low sample size data with many features,
methods in the field of multivariate statistical analysis are commonly
deployed. Here the aim is to find substructure in the data, or a simple
representation of the multi-dimensional space. Methods, such as Multiple
Correspondence Analysis (Le Roux and Rouanet 2010) are appropriate for
high dimensional data comprised of binary indicators. This technique is
cited in many studies in the application of dimension reduction and
clustering of comorbidities and phenotypes from disease (Han, Benseler,
and Tyrrell 2018; Costa et al. 2013).

In terms of patient survival, methods in statistical survival analysis
are well suited and commonly used in the field of genomics (Chen, Sun,
and Hoshida 2014). These methods in particular provide a mechanism for
dealing with right-censoring, a common feature of clinical studies where
the clinical outcome of interest may not be known by the end of the
study period.

An open research question exists around the clinical pathway and
survival time of children exhibiting bi-allelic loss-of-function
variants in \emph{SMPD4}. With current research highlighting a diverse
and heterogeneous phenotype, the relationship between these diverse
features is not well understood. Furthermore, the survival time of
affected children has not been extensively analysed despite early-demise
featuring as a severe outcome. While a connection has been highlighted
between some missense variants and a milder presentation with longer
survival (Magini et al. 2019), this hypothesis has not been analysed
further.

@@

The data contained in Magini et al. (2019) ``Summary of SMPD4-Related
Clinical Phenotype'' is an excel spreadsheet tabulating each of the 21
individuals from the study with clinical details recorded. The file has
21 columns (one for each individual), and 64 rows (one for each
phenotype or clinical remark).

The data were read into \texttt{R} (R Core Team 2022) without any
changes, so as to preserve the reproducibility of the data
transformation steps. The data were transposed to ensure it could be
presented as \emph{tidy} formatted data (Wickham 2014). This requires
that each variable (clinical phenotype) forms a column; each observation
(individual) forms a row and each type of observational unit forms a
table (every cell has just one item). This resulted in a long-formatted
data set of 21 rows (individuals) and 64 columns (clinical features). In
many cases, key clinical information was entered as free-text
descriptions, which rendered any attempt of meaningful analysis
impractical (Table~\ref{tbl-nontidy}). In these cases, the text was
tokenised by separating the list of clinical observations at each comma
and forming a binary indicator column noting its presence `1' or absence
`0' (Table~\ref{tbl-tidy}).

\hypertarget{tbl-nontidy}{}
\begin{longtable}{ll}
\caption{\label{tbl-nontidy}Example of non-tidy free-text descriptions of features. }\tabularnewline

\toprule
Family 1- Individual 1 & Family 1- Individual 4  \\ 
\midrule
short palpebral fissures, large ears, simple helices & short palpebral fissures, receding forehead \\ 
\bottomrule
\end{longtable}

\hypertarget{tbl-tidy}{}
\begin{longtable}{lrrrr}
\caption{\label{tbl-tidy}Example of tidy formatted data where individuals are transposed into
rows and text into binary indicators }\tabularnewline

\toprule
id & short palpebral fissures & large ears & simple helices & receding forehead \\ 
\midrule
Family 1- Individual 1 & 1 & 1 & 1 & 0 \\ 
Family 1- Individual 4 & 1 & 0 & 0 & 1 \\ 
\bottomrule
\end{longtable}

@@

\hypertarget{tbl-km}{}
\begin{longtable}{rrrrrrrr}
\caption{\label{tbl-km}Survival probabilities derived from Kaplan-Meier estimation for the
range of recorded survival times, including 95\% Confidence Interval. }\tabularnewline
